Using Random Walk Simulations to Calculate Ground State Energies in Quantum Physics

### 2020-04-20 Paper/main.tex
Spell-check and grammar corrections

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  In this paper, we present a novel way of calculating ground state energies for
  a quantum simulation using a Random Walk Monte Carlo simulation. The random
  walk consists of a walker making steps in random directions, with each step
  having a probability of causing the walker to die and the random walk to end,
  with the probability determined by a function of position. The function can be
  chosen to model specific quantum systems. The rate of death of the walkers is
  used to calculate the ground state energy. This paper explores the efficacy of
  such a method in reproducing the known ground state energy of an infinite
  square well, and the extensibility of the method to more complex
  systems. Using the infinite square well example, the simulation found the
  energy proportionality to be $1.2339 \pm 0.0008$, which is consister to high
  precision with the known $\frac{pi^2}{8}$ theoretical constant. Thus, we found
  that this method is very good at reproducing the known results to a high
  precision, and with good reliability, having very small variance in
  repeats. The method is extensible to higher dimensions, with a linear increase
  in time-complexity for each added dimension.
\end{abstract}

\section{Introduction}
\IEEEPARstart{T}{he} developments in computer power in recent years have opened
up the possibility for novel computational solutions to many problems that were
otherwise limited. We now have access to computers that can numerically solve
problems and run simulations that were previously too computationally intensive
to do.  One such problem is the calculation of the ground state energies of
different quantum systems. Many methods exist for solving these problems, using
both classical and quantum computational approaches. Whilst generally it may be
true that Quantum computating approahes are more suited to solving such problems
\cite{Mazzola}, these approaches are still in infancy and are not readily
accessible. Thus, there exist an abundance of classical computing approaches,
such as the Variational-Relaxiation algorithm \cite{Schroeder2017}. This is a
particularly useful algorithm for two-dimensional systems with non-separable
potentials but quickly becomes difficult to implement for more complex systems.

@@

Random walks descibe motion as a series of steps in some vector space, with the
direction of each step being random, i.e. with there being an assigned likelihood
of a step in each possible direction. Consider a simple 1-dimensional line and a
walker standing at a point labelled the origin. The walker takes 20 steps, the
direction of each step (forwards or backwards on the line) is random. The
position of the walker at each step is shown in Figure \ref{fig:cartoon}.

@@

\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker cannot die inside the well but
wil immediately die when they step outside, which we expect will mostly
happen given sufficient time. Inside the well, the potential is simply that of a
free particle \cite{Bransden2003}:

@@

\begin{equation}
  E = \lambda R^2 \frac{\hbar^2}{mJ^2}
  \label{eq:kasia}
\end{equation}
where only $\lambda$ is an unknown quantity. Thus, by calculating rate of death
$\lambda$, we can calculate the ground state energy of the infinite square
well. The known ground state energy of an infnite square well is
$\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ \cite{Davies1998}, so we expect to find $\lambda
R^2$ close to $\frac{\pi^2}{8}$.

@@

We first investigated an infinite square well with boundary $J=20$ and 1000000
walkers. We found the energy constant to be $1.2339 \pm 0.0008$, calculated from
the probability distribution of survival at a particular step. The probability
that a walker will survive at a particular step is simply the 1 minus the
normalised cumulative sum of walkers dying at each step. Figure
\ref{fig:cumplots} presents this probability and confirms the Anstatz of the
exponential function. The curve has been fitted with an exponential function of
the form in Equation \ref{eq:firsteq}. The gradient of Figure
\ref{fig:cumloghist} allowed for calculation of death rate which from Equation
\ref{eq:kasia} gives the calculated energy.

@@

As we can observe in Figure \ref{fig:cumhist}, the peak of the curve does not
lie at the origin as we might expect from the fact that all walkers should
survive at least the first step. The reason for this is twofold:
\begin{itemize}
\item For a well with boundary $J$, it is not possible for the walker to die in
  a number of steps less than $J$
\item It is highly unlikely that the walker will die at a number of steps
  greater than but still close to $J$ since it would require walking in the same
  direction a dispropotionate number of times.
\end{itemize}

\begin{figure}[H]%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/exp_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$. As can be seen, the peak is away from the origin. This
      is because, for a well with boundary $J$, the minimum number of steps a
      walker must take is $J$. However, it is still unlikely that the walker
      will die on this minimum number of steps, as then the walker must walk in
      the same direction every time. Thus, the most probable step for the walker
      to die on is greater than $J$. It is for this reason that the exponential
      decay does not begin at the origin in Figure \ref{fig:cumhist}, meaning
      some earlier steps must be neglected in order to achieve a better fit.}
    \label{fig:hist}
  \end{center}
\end{figure}

@@

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/line_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$ as in Figure \ref{fig:hist}, plotted on a logarithmic
      scale. As explained for Figure \ref{fig:hist}, the peak probability is on
      a step greater than $J$. This is clear in this plot. It is also evident
      that there is significantly more spread in the points at higher
      steps. This is due to there being significantly fewer walkers dying on
      these steps, so the relative differences are higher. This explains the
      fluctuations seen at higher steps in Figure \ref{fig:cumloghist}.}
    \label{fig:loghist}
  \end{center}
\end{figure}

@@

\begin{figure*}[ht!]
  \centering
  \subfloat[Probability of dying on a particlar step in Linear scale]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot_cutoff.pdf}
  \label{fig:expcutoff}}
  \centering
  \subfloat[Probability of surviving to at least a particular step in a
  logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot_cutoff.pdf}
  \label{fig:cum_line_cutoff_plot}}
  \caption{Histograms showing the results of the simulation when a maximm number
  of steps is introduced. If a walker reaches this number of steps, they
  automatically die, regardless of position. This results in a
  disproportionately high number of walkers dying at this maximum number, 2000
  in this case, as shown by Figure \ref{fig:expcutoff}. Although this presents
  no fluctuation at higher number of steps, as shown in Figure
  \ref{fig:cum_line_cutoff_plot}, some systematic error is introduced
  through this method.}
  \label{fig:cutoffplots}
\end{figure*}

@@

It is also necessary to investigate how varying the boundary $J$ affects the
accuracy of the final answer. Figure \ref{fig:multi_line_plot} presents the
probability plot from Figure \ref{fig:cumloghist} for many different boundary
$J$ values. No step cutoff was introduced. The calculated $\lambda J^2$ value is
shown in the legend. The gradient evidently changes, since the expected highest
number of steps would increase as the boundary is larger, as there is more
space for the walker to move before reaching a boundary. As shown in the legend,
the calculated $\lambda J^2$ values get closer and closer to the
theoretical value as $J$ increases. This is due to the fact that as $J$
increases, there are more walkers surviving to larger number of steps. Thus, we
can say that using a larger value of $J$ will result in a more accurate answer
but will also be more computationally intensive and take longer. Thus, we make a
tradeoff between speed and accuracy and choose $J=20$ for the final
investigation.

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random walk computational simulations. The
investigation focussed on a one dimensional infinite square well, but can easily
be extended to more complex systems such as a cirular well in two dimensions.

@@

\pagebreak
\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Consider a one-dimensional lattice upon which a walker is undergoing a random
walker. Let us denote the probability that the walker is at position $j$ after
$n$ steps as $p(j,n)$ and the probabilty that the walker will die at position
$j$ as $a(j)$. We will denote the position-step number coordinates as $[j,
n]$. Thus, for a walker to have coordinates $[j, n+1]$, in the previous step,
they must have had coordinates $[j \pm 1, n]$. So we can say:

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  In this paper, we present a novel way of calculating ground state energies for
  a quantum simulation using a Random Walk Monte Carlo simulation. The random
  walk consists of a walker making steps in random directions, with each step
  having a probability of causing the walker to die and the random walk to end,
  with the probability determined by a function of position. The function can be
  chosen to model specific quantum systems. The rate of death of the walkers is
  used to calculate the ground state energy. This paper explores the efficacy of
  such a method in reproducing the known ground state energy of an infinite
  square well, and the extensibility of the method to more complex
  systems. Using the infinite square well example, the simulation found the
  energy proportionality to be $1.2339 \pm 0.0008$, which is consistent to high
  precision with the known $\frac{\pi^2}{8}$ theoretical constant. Thus, we found
  that this method is very good at reproducing the known results to a high
  precision, and with good reliability, having very small variance in
  repeats. The method is extensible to higher dimensions, with a linear increase
  in time-complexity for each added dimension.
\end{abstract}

\section{Introduction}
\IEEEPARstart{T}{he} developments in computer power in recent years have opened
up the possibility for novel computational solutions to many problems that were
otherwise limited. We now have access to computers that can numerically solve
problems and run simulations that were previously too computationally intensive
to do.  One such problem is the calculation of the ground state energies of
different quantum systems. Many methods exist for solving these problems, using
both classical and quantum computational approaches. Whilst generally it may be
true that Quantum computing approaches are more suited to solving such problems
\cite{Mazzola}, these approaches are still in infancy and are not readily
accessible. Thus, there exist an abundance of classical computing approaches,
such as the Variational-Relaxation algorithm \cite{Schroeder2017}. This is a
particularly useful algorithm for two-dimensional systems with non-separable
potentials but quickly becomes difficult to implement for more complex systems.

@@

Random walks describe motion as a series of steps in some vector space, with the
direction of each step being random, i.e. with there being an assigned likelihood
of a step in each possible direction. Consider a simple 1-dimensional line and a
walker standing at a point labelled the origin. The walker takes 20 steps, the
direction of each step (forwards or backwards on the line) is random. The
position of the walker at each step is shown in Figure \ref{fig:cartoon}.

@@

\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker cannot die inside the well but
will immediately die when they step outside, which we expect will mostly
happen given sufficient time. Inside the well, the potential is simply that of a
free particle \cite{Bransden2003}:

@@

\begin{equation}
  E = \lambda R^2 \frac{\hbar^2}{mJ^2}
  \label{eq:kasia}
\end{equation}
where only $\lambda$ is an unknown quantity. Thus, by calculating rate of death
$\lambda$, we can calculate the ground state energy of the infinite square
well. The known ground state energy of an infinite square well is
$\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ \cite{Davies1998}, so we expect to find $\lambda
R^2$ close to $\frac{\pi^2}{8}$.

@@

We first investigated an infinite square well with boundary $J=20$ and 1000000
walkers. We found the energy constant to be $1.2339 \pm 0.0008$, calculated from
the probability distribution of survival at a particular step. The probability
that a walker will survive at a particular step is simply the 1 minus the
normalised cumulative sum of walkers dying at each step. Figure
\ref{fig:cumplots} presents this probability and confirms the ansatz of the
exponential function. The curve has been fitted with an exponential function of
the form in Equation \ref{eq:firsteq}. The gradient of Figure
\ref{fig:cumloghist} allowed for calculation of death rate which from Equation
\ref{eq:kasia} gives the calculated energy.

@@

As we can observe in Figure \ref{fig:cumhist}, the peak of the curve does not
lie at the origin as we might expect from the fact that all walkers should
survive at least the first step. The reason for this is twofold:
\begin{itemize}
\item For a well with boundary $J$, it is not possible for the walker to die in
  a number of steps less than $J$
\item It is highly unlikely that the walker will die at a number of steps
  greater than but still close to $J$ since it would require walking in the same
  direction a disproportionate number of times.
\end{itemize}

\begin{figure}[H]%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/exp_plot.pdf}
    \caption{The probability of dying on a particular step in an infinite square
      well with $J=20$. As can be seen, the peak is away from the origin. This
      is because, for a well with boundary $J$, the minimum number of steps a
      walker must take is $J$. However, it is still unlikely that the walker
      will die on this minimum number of steps, as then the walker must walk in
      the same direction every time. Thus, the most probable step for the walker
      to die on is greater than $J$. It is for this reason that the exponential
      decay does not begin at the origin in Figure \ref{fig:cumhist}, meaning
      some earlier steps must be neglected in order to achieve a better fit.}
    \label{fig:hist}
  \end{center}
\end{figure}

@@

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/line_plot.pdf}
    \caption{The probability of dying on a particular step in an infinite square
      well with $J=20$ as in Figure \ref{fig:hist}, plotted on a logarithmic
      scale. As explained for Figure \ref{fig:hist}, the peak probability is on
      a step greater than $J$. This is clear in this plot. It is also evident
      that there is significantly more spread in the points at higher
      steps. This is due to there being significantly fewer walkers dying on
      these steps, so the relative differences are higher. This explains the
      fluctuations seen at higher steps in Figure \ref{fig:cumloghist}.}
    \label{fig:loghist}
  \end{center}
\end{figure}

@@

\begin{figure*}[ht!]
  \centering
  \subfloat[Probability of dying on a particlar step in Linear scale]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot_cutoff.pdf}
  \label{fig:expcutoff}}
  \centering
  \subfloat[Probability of surviving to at least a particular step in a
  logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot_cutoff.pdf}
  \label{fig:cum_line_cutoff_plot}}
  \caption{Histograms showing the results of the simulation when a maximum number
  of steps is introduced. If a walker reaches this number of steps, they
  automatically die, regardless of position. This results in a
  disproportionately high number of walkers dying at this maximum number, 2000
  in this case, as shown by Figure \ref{fig:expcutoff}. Although this presents
  no fluctuation at higher number of steps, as shown in Figure
  \ref{fig:cum_line_cutoff_plot}, some systematic error is introduced
  through this method.}
  \label{fig:cutoffplots}
\end{figure*}

@@

It is also necessary to investigate how varying the boundary $J$ affects the
accuracy of the final answer. Figure \ref{fig:multi_line_plot} presents the
probability plot from Figure \ref{fig:cumloghist} for many different boundary
$J$ values. No step cutoff was introduced. The calculated $\lambda J^2$ value is
shown in the legend. The gradient evidently changes, since the expected highest
number of steps would increase as the boundary is larger, as there is more
space for the walker to move before reaching a boundary. As shown in the legend,
the calculated $\lambda J^2$ values get closer and closer to the
theoretical value as $J$ increases. This is due to the fact that as $J$
increases, there are more walkers surviving to larger number of steps. Thus, we
can say that using a larger value of $J$ will result in a more accurate answer
but will also be more computationally intensive and take longer. Thus, we make a
trade-off between speed and accuracy and choose $J=20$ for the final
investigation.

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random walk computational simulations. The
investigation focussed on a one dimensional infinite square well, but can easily
be extended to more complex systems such as a circular well in two dimensions.

@@

\pagebreak
\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Consider a one-dimensional lattice upon which a walker is undergoing a random
walker. Let us denote the probability that the walker is at position $j$ after
$n$ steps as $p(j,n)$ and the probability that the walker will die at position
$j$ as $a(j)$. We will denote the position-step number coordinates as $[j,
n]$. Thus, for a walker to have coordinates $[j, n+1]$, in the previous step,
they must have had coordinates $[j \pm 1, n]$. So we can say:
